Code used to generate paper results : CT characteristics

### 2024-07-02 Paper_code-CT_analysis.Rmd
Updated with new tumors thresholds
Adapted all code to latest version of paper

---
title: "Code used to generate paper results : Database comparison and CT characteristics"
author: "Julie Devis"
date: "2023-08-28"
output: 
  html_document:
    theme: cosmo
    df_print : paged
    highlight : default
    toc: yes
    number_sections : TRUE
    toc_depth : 4
---

@@

```{r library, message = FALSE}
library(readxl)
library(readr)
library(CTexploreR)
library(tidyverse)
library(SummarizedExperiment)
library(ComplexHeatmap)
library(circlize)
library(SingleCellExperiment)
library(msigdbr)
library(DOSE)
library(org.Hs.eg.db)
library(clusterProfiler)
library(patchwork)
library(AUCell)
```

```{r load_data, warning=FALSE, message=FALSE}
load("../CTdata/eh_data/all_genes.rda")
load("../CTdata/eh_data/GTEX_data.rda")
load("../CTdata/eh_data/CCLE_data.rda")
load("../CTdata/eh_data/TCGA_TPM.rda")
load("../CTdata/eh_data/normal_tissues_multimapping_data.rda")
load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
load("../CTdata/eh_data/DAC_treated_cells_multimapping.rda")
load("../CTdata/eh_data/DAC_treated_cells.rda")
load("../CTdata/eh_data/testis_sce.rda")
load("../CTdata/eh_data/FGC_sce.rda")
load("../CTdata/eh_data/oocytes_sce.rda")

load("../CTdata/eh_data/CT_genes.rda")

CT_genes_types <- CT_genes %>% 
  dplyr::rename(gene = external_gene_name) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ 
                                  "Not methylation")) %>%
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) %>% 
  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
                             !regulated_by_methylation & X_linked ~ "X_not_met",
                             regulated_by_methylation & !X_linked ~ "not_X_met",
                             !regulated_by_methylation & !X_linked ~ 
                               "not_X_not_met")) %>% 
  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ 
                                   "Testis specific",
                                 GTEX_category == "testis_preferential" ~ 
                                   "Testis preferential",
                                 GTEX_category == "lowly_expressed" ~ 
                                   "Undetectable in GTEX")) %>% 
  mutate(specificity = factor(specificity,
                              levels = c("Testis specific",  
                                         "Undetectable in GTEX", 
                                         "Testis preferential"))) %>% 
  mutate(chr_met = factor(chr_met, 
                          levels = c("X_met", "not_X_met", "X_not_met", 
                                     "not_X_not_met")))

all_genes_types <- all_genes %>% 
  dplyr::rename(gene = external_gene_name) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ 
                                  "Not methylation")) %>%
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) %>% 
  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
                             !regulated_by_methylation & X_linked ~ "X_not_met",
                             regulated_by_methylation & !X_linked ~ "not_X_met",
                             !regulated_by_methylation & !X_linked ~ 
                               "not_X_not_met")) %>% 
  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ 
                                   "Testis specific",
                                 GTEX_category == "testis_preferential" ~ 
                                   "Testis preferential",
                                 GTEX_category == "lowly_expressed" ~ 
                                   "Undetectable in GTEX")) %>% 
  mutate(specificity = factor(specificity,
                              levels = c("Testis specific",  
                                         "Undetectable in GTEX", 
                                         "Testis preferential"))) %>% 
  mutate(chr_met = factor(chr_met, 
                          levels = c("X_met", "not_X_met", "X_not_met", 
                                     "not_X_not_met")))

@@

Chr_colors = c("X-linked" = "deeppink3", "not X" = "lightgreen")

@@

@@

table(all_genes$CT_gene_type) #Before and after TSS verif
```

@@

```{r GTEX_function}
print_GTEX_heatmap <- function(genes = NULL, units = "log_TPM", 
                               row_fontsize = 2,
                               column_fontsize = 6, 
                               h_width = 3,
                               h_height = 10,
                               split_rows_by = "regulation") {

@@

  gene_type <- all_genes_types %>% 
    filter(gene %in% genes)

   if (split_rows_by == "regulation") {
    gene_type <- gene_type %>% arrange(Regulation)
    rowSplit <- gene_type$Regulation
  }
  if (split_rows_by == "testis_specificity") {
    gene_type <- gene_type %>% arrange(Testis_specificity)
    rowSplit <- gene_type$Testis_specificity
  }
  if (split_rows_by == "chr_met") {
    gene_type <- gene_type %>% arrange(chr_met)
    rowSplit <- gene_type$chr_met
  }
  if (split_rows_by == "specificity") {
    gene_type <- gene_type %>% arrange(specificity)
    rowSplit <- gene_type$specificity
  }

  if (split_rows_by == "GTEX_category") {
    GTEX_category <- gene_type %>% arrange(GTEX_category)
    rowSplit <- gene_type$GTEX_category
  }

@@

  specificity <- as.matrix(gene_type$specificity)
  rownames(specificity) <- gene_type$gene
  row_ha_ts <- rowAnnotation(specificity = specificity,
                             annotation_legend_param = legends_param,
                             simple_anno_size = unit(0.2, "cm"),
                             col = list(specificity = 
                                          Testis_specificity_colors),
                             annotation_name_gp = gpar(fontsize = 0))

@@

  h <- Heatmap(mat[gene_type$gene, ],
               name = "log(TPM +1)",
               column_title = "Normal tissues (GTEx)",
               column_title_gp = gpar(fontsize = 6, face = "bold"),
               row_split = rowSplit,
               cluster_row_slices = FALSE,
               row_title_gp = gpar(fontsize = 4, face = "bold"),
               border = TRUE,
               border_gp = gpar(lwd = 0.5),
               col = heatmap_colors,
               cluster_rows = TRUE,
               cluster_columns = TRUE,
               show_row_dend = FALSE,
               show_column_dend = FALSE,
               row_names_gp = gpar(fontsize = row_fontsize),
               column_names_gp = gpar(fontsize = column_fontsize),
               column_names_side = "top",
               # right_annotation = c(row_ha_regulation, 
               #                        row_ha_chr, 
               #                        row_ha_ts,
               #                        gap = unit(1, "mm")),
               row_names_side = "left",
               width = unit(h_width, "cm"),
               height = unit(h_height, "cm"),
               clustering_method_rows = "ward.D",
               use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10,
               heatmap_legend_param = legends_param)

@@

```{r TCGA_function}
print_TCGA_heatmap <- function(genes = NULL,
                               tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
                                            "TCGA-LUSC", "TCGA-COAD", 
                                            "TCGA-BRCA", "TCGA-HNSC",
                                            "TCGA-ESCA"),
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               split_rows_by = "regulation", 
                               h_width = 5, 
                               h_height = 10,
                               clustRow = TRUE,
                               clustCol = TRUE) {

@@

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(TCGA)
  rownames(mat) <- rowData(TCGA)$external_gene_name
  mat <- mat[rownames(mat) %in% all_genes$external_gene_name,]
  mat <- log1p(mat)

  gene_type <- all_genes_types %>% 
    filter(gene %in% genes)

  rowSplit <- NULL
  if(!is.null(split_rows_by)){
   if (split_rows_by == "regulation") {
    gene_type <- gene_type %>% arrange(Regulation)
    rowSplit <- gene_type$Regulation
  }
  if (split_rows_by == "testis_specificity") {
    gene_type <- gene_type %>% arrange(Testis_specificity)
    rowSplit <- gene_type$testis_specificity
  }
  if (split_rows_by == "specificity") {
    gene_type <- gene_type %>% arrange(specificity)
    rowSplit <- gene_type$specificity
  }

  if (split_rows_by == "chr_met") {
    gene_type <- gene_type %>% arrange(chr_met)
    rowSplit <- gene_type$chr_met
  }

  if (split_rows_by == "GTEX_category") {
    GTEX_category <- gene_type %>% arrange(GTEX_category)
    rowSplit <- gene_type$GTEX_category
  }
  }

@@

  specificity <- as.matrix(gene_type$specificity)
  rownames(specificity) <- gene_type$gene
  row_ha_ts <- rowAnnotation(specificity = specificity,
                             annotation_legend_param = legends_param,
                             simple_anno_size = unit(0.2, "cm"),
                             col = list(specificity = 
                                          Testis_specificity_colors),
                             annotation_name_gp = gpar(fontsize = 0))

@@

  Heatmap(mat[gene_type$gene, ],
          name = "TCGA (logTPM)",
          column_title = "TCGA tumor samples",
          column_title_gp = gpar(fontsize = 6, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          show_row_names = FALSE,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = 0),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          top_annotation = column_ha_tumor,
          right_annotation = c(row_ha_ts,
                               #row_ha_regulation,
                               #row_ha_chr,
                               gap = unit(1, "mm")),
          row_split = rowSplit,
          cluster_row_slices = FALSE,
          width = unit(h_width, "cm"),
          height = unit(h_height, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = FALSE)
}

@@

```{r CCLE_function}
print_CCLE_heatmap <- function(genes = NULL,
                               tum_type = c("Skin", "Lung", 
                                            "Colorectal", 
                                            "Breast", 
                                            "Head_and_Neck",
                                            "Esophageal"),
                               h_width = 5,
                               h_height = 10,
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               split_rows_by = "regulation", 
                               clustRow = TRUE,
                               clustCol = TRUE){

@@

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(CCLE)
  rownames(mat) <- rowData(CCLE)$external_gene_name
  mat <- mat[rownames(mat) %in% all_genes$external_gene_name, ]
  mat <- log1p(mat)

  gene_type <- all_genes_types %>% 
    filter(gene %in% genes)

   if (split_rows_by == "regulation") {
    gene_type <- gene_type %>% arrange(Regulation)
    rowSplit <- gene_type$Regulation
  }
  if (split_rows_by == "testis_specificity") {
    gene_type <- gene_type %>% arrange(testis_specificity)
    rowSplit <- gene_type$testis_specificity
  }
  if (split_rows_by == "specificity") {
    gene_type <- gene_type %>% arrange(specificity)
    rowSplit <- gene_type$specificity
  }
  if (split_rows_by == "chr_met") {
    gene_type <- gene_type %>% arrange(chr_met)
    rowSplit <- gene_type$chr_met
  }

   if (split_rows_by == "GTEX_category") {
    GTEX_category <- gene_type %>% arrange(GTEX_category)
    rowSplit <- gene_type$GTEX_category
  }

@@

  Heatmap(mat[gene_type$gene, ],
          name = "CCLE (logTPM)",
          column_title = paste0("CCLE tumor cell lines"),
          column_title_gp = gpar(fontsize = 6, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          show_row_names = TRUE,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = fontsize),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          top_annotation = column_ha_tumor,
          # right_annotation = c(row_ha_ts,
          #                      row_ha_regulation,
          #                      #row_ha_chr,
          #                      gap = unit(1, "mm")),
          row_split = rowSplit,
          width = unit(h_width, "cm"),
          height = unit(h_width, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = FALSE)
}
```

@@

```{r selection_heatmap}
my_genes <- CT_genes %>% 
  pull(external_gene_name)

ht_TCGA <- print_TCGA_heatmap(my_genes, 
                              fontsize = 4, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
                                           #"TCGA-ESCA", "TCGA-LUSC", 
                                           #"TCGA-COAD",
                                           #"TCGA-BRCA",
                                           "TCGA-HNSC"),
                              clust_method = "ward.D",
                              split_rows_by = "specificity")

ht_CCLE <- print_CCLE_heatmap(my_genes, 
                              downsample_to_n = 50, 
                              tum_type = c("Skin", "Lung", 
                                           #"colorectal", 
                                           #"breast", 
                                           #"esophageal",
                                           "Head_and_Neck"),
                              #"Leukemia", "Lymphoma"),
                              h_width = 3,
                              h_height = 8,
                              split_rows_by = "specificity")

ht_GTEX <- print_GTEX_heatmap(my_genes, 
                              row_fontsize = 1,
                              column_fontsize = 5,
                              h_width = 5,
                              h_height = 8,
                             split_rows_by = "specificity")
h_list <- ht_GTEX + ht_CCLE + ht_TCGA
draw(h_list, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01), "cm"),
     merge_legend = TRUE)
```

```{r, eval = FALSE, echo=FALSE}
pdf(file = paste0("figs/GTEX_Tumors.pdf"))
draw(h_list, main_heatmap = "TCGA (logTPM)", 
     ht_gap = unit(c(0.3, 0.01), "cm"),
     merge_legend = TRUE)
dev.off()
```

Counter examples

```{r counter_examples}
genes_not_testis <- all_genes %>% 
  filter(GTEX_category != "lowly_expressed") %>% 
  filter(testis_specificity == "not_testis_specific") %>% 
  pull(external_gene_name) %>% 
  sample(2)

genes_not_TCGA_activated <- all_genes %>% 
  filter(GTEX_category != "lowly_expressed") %>% 
  filter(testis_specificity != "not_testis_specific") %>% 
  filter(!TCGA_category %in% c("activated", "multimapping_issue")) %>% 
  pull(external_gene_name) %>% 
  sample(1)

genes_not_ccle_activated <- all_genes %>% 
  filter(GTEX_category != "lowly_expressed") %>% 
  filter(testis_specificity != "not_testis_specific") %>% 
  filter(CCLE_category %in% c("activated")) %>% 
  pull(external_gene_name) %>% 
  sample(1)

my_genes_counter <- c(genes_not_testis, genes_not_TCGA_activated, 
                      genes_not_ccle_activated)

ht_TCGA_counter <- print_TCGA_heatmap(my_genes_counter, 
                              fontsize = 4, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD",
                                           "TCGA-HNSC"),
                              clust_method = "ward.D",
                              split_rows_by = "specificity")

>>>>

---
title: "Code used to generate paper results : CT characteristics"
author: "Julie Devis"
date: "2023-08-28"
output: 
  html_document:
    theme: cosmo
    df_print : paged
    highlight : default
    toc: yes
    number_sections : TRUE
    toc_depth : 4
---

@@

```{r library, message = FALSE}
library(readxl)
library(readr)
library(CTexploreR)
library(tidyverse)
library(SummarizedExperiment)
library(ComplexHeatmap)
library(circlize)
library(SingleCellExperiment)
library(msigdbr)
library(DOSE)
library(org.Hs.eg.db)
library(clusterProfiler)
library(patchwork)
library(Vennerable)
```

```{r load_data, warning=FALSE, message=FALSE}
load("../CTdata/eh_data/all_genes.rda")
load("../CTdata/eh_data/GTEX_data.rda")
load("../CTdata/eh_data/CCLE_data.rda")
load("../CTdata/eh_data/TCGA_TPM.rda")
load("../CTdata/eh_data/normal_tissues_multimapping_data.rda")
load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
load("../CTdata/eh_data/DAC_treated_cells_multimapping.rda")
load("../CTdata/eh_data/DAC_treated_cells.rda")
load("../CTdata/eh_data/testis_sce.rda")
load("../CTdata/eh_data/FGC_sce.rda")
load("../CTdata/eh_data/oocytes_sce.rda")
load("../CTdata/eh_data/CT_genes.rda")

load("~/cluster/DataSets/FGC/scWGBS/processed_data/scWGBS_FGC_SE.rda")  
load("~/cluster/CBIOProjects/202406_CT_genes_paper/tmp_data/CT_promoter_regions_hg19_GR.rda")

all_genes_clean <- all_genes %>% 
  dplyr::rename(gene = external_gene_name) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ 
                                  "Not methylation")) %>%
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) %>% 
  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
                             !regulated_by_methylation & X_linked ~ "X_not_met",
                             regulated_by_methylation & !X_linked ~ "not_X_met",
                             !regulated_by_methylation & !X_linked ~ 
                               "not_X_not_met")) %>% 
  mutate(GTEX_specificity = case_when(GTEX_category == "testis_specific" ~ 
                                        "Testis specific",
                                        GTEX_category == "testis_preferential" ~ 
                                        "Testis preferential",
                                        GTEX_category == "lowly_expressed" ~ 
                                        "Undetectable in GTEX")) %>% 
  mutate(GTEX_specificity = factor(GTEX_specificity, 
                                     levels = c("Testis specific",  
                                                "Undetectable in GTEX", 
                                                "Testis preferential"))) %>% 
  mutate(testis_specificity = case_when(testis_specificity == 
                                          "testis_specific" ~ "Testis specific",
                                        testis_specificity == 
                                          "testis_preferential" ~ 
                                          "Testis preferential")) %>%
  mutate(testis_specificity = factor(testis_specificity, 
                                     levels = c("Testis specific", 
                                                "Testis preferential"))) %>% 
  mutate(chr_met = factor(chr_met, 
                          levels = c("X_met", "not_X_met", "X_not_met", 
                                     "not_X_not_met"))) %>% 
  mutate(met_3_groups = case_when(regulated_by_methylation & X_linked ~ 
                                    "Methylation X-linked",
                                   regulated_by_methylation & !X_linked ~ 
                                    "Methylation not X-linked",
                             !regulated_by_methylation ~ "Not Methylation")) %>% 
  mutate(met_3_groups = factor(met_3_groups, 
                          levels = c("Methylation X-linked", 
                                     "Methylation not X-linked", 
                                     "Not Methylation")))

CT_genes <- all_genes %>%
  filter(CT_gene_type == "CT_gene")
CT_genes_types <- all_genes_clean %>% 
  filter(CT_gene_type == "CT_gene")

CTP_genes <- all_genes %>%
  filter(CT_gene_type == "CTP_gene")
CTP_genes_types <- all_genes_clean %>% 
  filter(CT_gene_type == "CTP_gene")
```

@@

Chr_colors = c("X-linked" = "deeppink3", "not X" = "lightgreen")
Chr_colors_with_Y = c("X-linked" = "deeppink3", "Autosome" = "lightgreen", 
                      "Y-linked" = "black")

GTEX_specificity_colors = c("Testis specific" = "mediumpurple4",
                            "Testis preferential" = "goldenrod3",
                            "Undetectable in GTEX" = "gray")

@@

Fetal_cellType_colors <- c( 
"F_GC" = "pink3",
  "F_oogonia" = "palevioletred3",
  "F_oocyte" = "mediumorchid4",
  "M_GC" = "steelblue3",
  "M_pre_spermatogonia" = "royalblue3",
  "F_soma" = "gray",
  "M_soma" = "gray")

Fetal_time_colors <- c("6W" = "aliceblue",
                      "7W" = "cyan",
                      "8W" = "cyan4",
                      "9W" = "skyblue1",
                      "10W" = "dodgerblue",
                      "17W" = "blue",
                      "21W" = "lightblue4",
                      "24W" = "black")

@@

table(all_genes$IGV_backbone)
table(all_genes$CT_gene_type)
```

@@

```{r GTEX_function}
print_GTEX_heatmap <- function(genes = NULL, units = "log_TPM", 
                               row_fontsize = 2,
                               column_fontsize = 6, 
                               h_width = 3,
                               h_height = 10,
                               split_rows_by = "regulation",
                               clust_method = "ward.D",
                               clustRow = FALSE) {

@@

  gene_type <- all_genes_clean %>% 
    filter(gene %in% genes) %>% 
    mutate(gene = factor(gene, levels = genes)) %>% 
    arrange(gene)

    if (split_rows_by == "regulation") {
        if (clustRow) gene_type <- gene_type %>% 
            arrange(Regulation)
      rowSplit <- gene_type$Regulation
    }
    if (split_rows_by == "testis_specificity") {
        if (clustRow) gene_type <- gene_type %>% 
            arrange(testis_specificity)
      rowSplit <- gene_type$testis_specificity
    }
    if (split_rows_by == "chr_met") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(chr_met)
      rowSplit <- gene_type$chr_met
    }
    if (split_rows_by == "GTEX_specificity") {
      if (clustRow) GTEX_category <- gene_type %>% 
          arrange(GTEX_specificity)
      rowSplit <- gene_type$GTEX_specificity
    }

@@

  GTEX_specificity <- as.matrix(gene_type$GTEX_specificity)
  rownames(GTEX_specificity) <- gene_type$gene
  row_ha_GTEX <- rowAnnotation(specificity = GTEX_specificity,
                             annotation_legend_param = legends_param,
                             simple_anno_size = unit(0.2, "cm"),
                             col = list(specificity = 
                                          GTEX_specificity_colors),
                             annotation_name_gp = gpar(fontsize = 0))

  testis_specificity <- as.matrix(gene_type$testis_specificity)
  rownames(testis_specificity) <- gene_type$gene
  row_ha_TS <- rowAnnotation(testis_specificity = testis_specificity,
                             annotation_legend_param = legends_param,
                             simple_anno_size = unit(0.2, "cm"),
                             col = list(testis_specificity = 
                                          Testis_specificity_colors),
                             annotation_name_gp = gpar(fontsize = 0))

@@

  h <- Heatmap(mat[as.vector(gene_type$gene), ],
               name = "log(TPM +1)",
               column_title = "Normal tissues (GTEx)",
               column_title_gp = gpar(fontsize = 6, face = "bold"),
               row_split = rowSplit,
               cluster_row_slices = FALSE,
               row_title_gp = gpar(fontsize = 6, face = "bold"),
               border = TRUE,
               border_gp = gpar(lwd = 0.5),
               col = heatmap_colors,
               cluster_rows = clustRow,
               cluster_columns = TRUE,
               show_row_dend = FALSE,
               show_column_dend = FALSE,
               row_names_gp = gpar(fontsize = row_fontsize),
               column_names_gp = gpar(fontsize = column_fontsize),
               column_names_side = "top",
               # right_annotation = c(row_ha_regulation, 
               #                        row_ha_chr, 
               #                        row_ha_ts,
               #                        gap = unit(1, "mm")),
               row_names_side = "left",
               width = unit(h_width, "cm"),
               height = unit(h_height, "cm"),
               clustering_method_rows = clust_method,
               use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10,
               heatmap_legend_param = legends_param)

@@

```{r TCGA_function}
print_TCGA_heatmap <- function(genes = NULL,
                               tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
                                            "TCGA-LUSC", "TCGA-COAD", 
                                            "TCGA-BRCA", "TCGA-HNSC",
                                            "TCGA-ESCA"),
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               split_rows_by = "regulation", 
                               h_width = 5, 
                               h_height = 10,
                               show_annotation_legend = TRUE,
                               show_top_annotations = TRUE,
                               show_row_names = FALSE,
                               clustRow = FALSE,
                               clustCol = TRUE) {

@@

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(TCGA)
  rownames(mat) <- rowData(TCGA)$external_gene_name
  mat <- log1p(mat)

  gene_type <- all_genes_clean %>% 
    filter(gene %in% genes) %>% 
    mutate(gene = factor(gene, levels = genes)) %>% 
    arrange(gene)

  rowSplit <- NULL
  if(!is.null(split_rows_by)){
    if (split_rows_by == "regulation") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(Regulation)
      rowSplit <- gene_type$Regulation
    }
    if (split_rows_by == "testis_specificity") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(testis_specificity)
      rowSplit <- gene_type$testis_specificity
    }
    if (split_rows_by == "GTEX_specificity") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(GTEX_specificity)
      rowSplit <- gene_type$GTEX_specificity
    }

    if (split_rows_by == "chr_met") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(chr_met)
      rowSplit <- gene_type$chr_met
    }
  }

@@

  GTEX_specificity <- as.matrix(gene_type$GTEX_specificity)
  rownames(GTEX_specificity) <- gene_type$gene
  row_ha_GTEX <- rowAnnotation(GTEX_specificity = GTEX_specificity,
                             annotation_legend_param = legends_param,
                             simple_anno_size = unit(0.2, "cm"),
                             col = list(GTEX_specificity = 
                                          GTEX_specificity_colors),
                             annotation_name_gp = gpar(fontsize = 0))

  testis_specificity <- as.matrix(gene_type$testis_specificity)
  rownames(testis_specificity) <- gene_type$gene
  row_ha_TS <- rowAnnotation(specificity = testis_specificity,
                             annotation_legend_param = legends_param,
                             simple_anno_size = unit(0.2, "cm"),
                             col = list(testis_specificity = 
                                          Testis_specificity_colors),
                             annotation_name_gp = gpar(fontsize = 0))

  rAnnot <- NULL
  if(show_annotation_legend){
    if (split_rows_by == "regulation") {
  rAnnot <- c(row_ha_regulation, gap = unit(1, "mm"))
    }
    if (split_rows_by == "testis_specificity") {
  rAnnot <- c(row_ha_TS, gap = unit(1, "mm"))
    }
    if (split_rows_by == "GTEX_specificity") {
  rAnnot <- c(row_ha_GTEX, gap = unit(1, "mm"))
    }

    if (split_rows_by == "chr_met") {
  rAnnot <- c(row_ha_TS, gap = unit(1, "mm"))
    }
  }

@@

  tAnnot <- NULL
  if (show_top_annotations) tAnnot <- column_ha_tumor

  Heatmap(mat[as.vector(gene_type$gene), ],
          name = "TCGA (logTPM)",
          column_title = "TCGA tumor samples",
          column_title_gp = gpar(fontsize = 6, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          show_row_names = show_row_names,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = fontsize),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          top_annotation = tAnnot,
          right_annotation = rAnnot,
          row_split = rowSplit,
          cluster_row_slices = FALSE,
          width = unit(h_width, "cm"),
          height = unit(h_height, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = FALSE)
}
```

@@

```{r CCLE_function}
print_CCLE_heatmap <- function(genes = NULL,
                               tum_type = c("Skin", "Lung", 
                                            "Colorectal", 
                                            "Breast", 
                                            "Head_and_Neck",
                                            "Esophageal"),
                               h_width = 5,
                               h_height = 10,
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               split_rows_by = "regulation", 
                               show_annotation_legend = TRUE,
                               clustRow = FALSE,
                               clustCol = TRUE){

@@

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(CCLE)
  rownames(mat) <- rowData(CCLE)$external_gene_name
  mat <- log1p(mat)

  gene_type <- all_genes_clean %>% 
    filter(gene %in% genes) %>% 
    mutate(gene = factor(gene, levels = genes)) %>% 
    arrange(gene)

 rowSplit <- NULL
  if(!is.null(split_rows_by)){
    if (split_rows_by == "regulation") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(Regulation)
      rowSplit <- gene_type$Regulation
    }
    if (split_rows_by == "testis_specificity") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(testis_specificity)
      rowSplit <- gene_type$testis_specificity
    }
    if (split_rows_by == "GTEX_specificity") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(GTEX_specificity)
      rowSplit <- gene_type$GTEX_specificity
    }

    if (split_rows_by == "chr_met") {
      if (clustRow) gene_type <- gene_type %>% 
          arrange(chr_met)
      rowSplit <- gene_type$chr_met
    }
  }

@@

  Heatmap(mat[as.vector(gene_type$gene), ],
          name = "CCLE (logTPM)",
          column_title = paste0("CCLE tumor cell lines"),
          column_title_gp = gpar(fontsize = 6, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          show_row_names = TRUE,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = fontsize),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          top_annotation = column_ha_tumor,
          right_annotation = rAnnot,
          row_split = rowSplit,
          width = unit(h_width, "cm"),
          height = unit(h_width, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = FALSE)
}
```

@@

```{r selection_heatmap}
my_genes <- all_genes %>% 
  filter(CT_gene_type != "other") %>% 
  pull(external_gene_name)

# Run the heatmap with custRow parameters to define the correct order
# This order will then be used for GTEX and CCLE heatmaps
ht_TCGA <- print_TCGA_heatmap(my_genes, 
                              fontsize = 4, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
                                           #"TCGA-ESCA", "TCGA-LUSC", 
                                           #"TCGA-COAD",
                                           #"TCGA-BRCA",
                                           "TCGA-HNSC"),
                              clust_method = "ward.D",
                              clustRow = TRUE,
                              split_rows_by = "testis_specificity")
indices <- row_order(ht_TCGA)
gene_order <- all_genes_clean %>%
  filter(gene %in% my_genes) %>%
  arrange(testis_specificity) %>%
  pull(gene)

TS_order <- gene_order[indices$`Testis specific`]
TP_order <- gene_order[indices$`Testis preferential`]
my_genes <- c(TS_order, TP_order)

ht_TCGA <- print_TCGA_heatmap(my_genes, 
                              fontsize = 4, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
                                           #"TCGA-ESCA", "TCGA-LUSC", 
                                           #"TCGA-COAD",
                                           #"TCGA-BRCA",
                                           "TCGA-HNSC"),
                              clust_method = "ward.D",
                              show_annotation_legend = FALSE,
                              clustRow = FALSE,
                              split_rows_by = "testis_specificity")

ht_CCLE <- print_CCLE_heatmap(my_genes, 
                              downsample_to_n = 50, 
                              tum_type = c("Skin", "Lung", 
                                           #"colorectal", 
                                           #"breast", 
                                           #"esophageal",
                                           "Head_and_Neck"),
                              #"Leukemia", "Lymphoma"),
                              clustRow = FALSE,
                              show_annotation_legend = FALSE,
                              h_width = 3,
                              h_height = 8,
                              split_rows_by = "testis_specificity")

ht_GTEX <- print_GTEX_heatmap(my_genes, 
                              row_fontsize = 1,
                              column_fontsize = 5,
                              h_width = 5,
                              h_height = 8,
                              clustRow = FALSE,
                              split_rows_by = "testis_specificity")

h_list <- ht_GTEX + ht_CCLE + ht_TCGA
draw(h_list, main_heatmap = "log(TPM +1)", ht_gap = unit(c(0.3, 0.01), "cm"),
     merge_legend = TRUE)
```

```{r, eval = FALSE, echo=FALSE}
pdf(file = paste0("figs/GTEX_Tumors.pdf"))
draw(h_list, main_heatmap = "log(TPM +1)", 
     ht_gap = unit(c(0.3, 0.01), "cm"),
     merge_legend = TRUE)
dev.off()
```
